Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-09-02 summary.Rmd
updated theorem 5 and added second example

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation"
author: John Koo, Minh Tang, Michael Trosset
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
fontsize: 12pt
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,vlined,ruled]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
bibliography: misc.bib
abstract: |
  In this paper, we connect two probabilistic models for graphs, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  clustering and parameter estimation. In particular, we note that the PABM 
  can be represented as latent positions such that points within the same 
  cluster lie on a subspace, and the subspaces that represent each 
  cluster are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding (ASE) of the GRDPG, we 
  are able to establish theoretical asymptotic results of our clustering and 
  parameter estimation methods for the PABM. 
---

@@

\newcommand{\diag}{\text{diag}}

@@

Statistical analysis on graphs and networks often involve the partitioning of a 
graph into clusters. This is often motivated by the belief that there is an 
underlying and unobserved community or cluster label for each vertex of the 
graph, and edges between pairs of vertices are determined by drawing from a 
probability distribution based on the communities assigned to each pair. 
One such probabilistic model is the 
Stochastic Block Model (SBM), which assumes that the edge probability from 
one vertex to another follows a Beronulli distribution with fixed probabilities 
for each pair of community labels. 
The Popularity Adjusted Block Model (PABM) was then introduced by 
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the SBM to 
address the heterogeneity of edge probabilities within and between communities 
while still maintaining the community structure.

The Random Dot Product Graph (RDPG) model \cite{athreya2017statistical} is 
another graph model with Beronulli edge probabilities. Under this model, each 
vertex of the graph can be represented by a point in latent space such that 
the edge probability between any pair of vertices is given by their 
corresponding dot product in the latent space. The SBM is equivalent to a 
special case of the RDPG model in which all vertices in a given community share 
the same position in the latent space. It has also been shown that similar 
probabilistic graph models, (such as the Mixed Membership Stochastic Block Model
\cite{rubindelanchy2017statistical}). An analogous property exists for the 
PABM not under the RDPG model but under the *Generalized* Random Dot Product 
Graph (GRDPG) model.

In this paper, we will only consider undirected graphs, that is the edge value 
from vertices $i$ and $j$ is equal to the edge in the opposite direction, from 
vertices $j$ and $i$. Furthermore, we will only consider unweighted graphs with 
binary edge values. We will also assume that graphs are hollow, i.e., there 
are no self-loops.

@@

## The popularity adjusted block model (PABM) \cite{noroozi2019estimation}

**Definition 1**. 
Let $P \in [0, 1]^{n \times n}$ be an edge probability matrix for a set of $n$ 
vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
$P^{(kl})$ describes the edge probabilities between vertices in communities 
$k$ and $l$. 
Let graph $G = (V, E)$ be an undirected, unweighted graph such 
that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
realization of $Bernoulli(P)$, i.e., 
$A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ for $i > j$ 
($A_{ij} = A_{ji}$ and $A_{ii} = 0$).

@@

## The generalized random dot product graph (GRDPG) \cite{rubindelanchy2017statistical}

**Definition 2**.
Let $X \in \mathbb{R}^{n \times d}$ be latent positions of the vertices of a 
graph $G$. $X$ consists of row vectors $x_i^\top$. Let 
$A \in \{0, 1\}^{n \times n}$ be the corresponding adjacency matrix.

Fix $p$, $q$ such that $p + q = d$ and define 
$I_{pq} = \begin{bmatrix} I_p & 0 \\ 0 & -I_q \end{bmatrix}$.

Then $G = (V, E)$ is a generalized random dot product graph with signature 
$(p, q)$ and patent positions $X$ iff its random adjacency matrix can be 
described as $A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ where 
$P_{ij} = x_i^\top I_{pq} x_j$.

@@

**Remark**. 
Like the RDPG, the latent positions of a GRDPG are not unique. More 
specifically, if $P_{ij} = x_i^\top I_{pq} x_j$, then we also have for any 
$Q \in \{R : R^\top I_{pq} R = I_{pq} \}$, 
$(Q x_i)^\top I_{pq} (Q x_j) = x_i^\top (Q^\top I_{pq} Q) x_j = 
x_i^\top I_{pq} x_j = P_{ij}$ (this is the indefinite orthogonal group, 
$\mathbb{O}(p, q)$).
Unlike in the RDPG case, transforming the latent positions by multiplication 
with $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or 
distances.

@@

* Positions 1, 5, 9 are fixed
* The cycles of order 2 are
    * $(2, 4)$
    * $(3, 7)$
    * $(6, 8)$

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation using the adjacency matrix.}
  \KwData{Adjacency matrix $A$, cluster assignments $1, ..., K$}
  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(kl)}\}_K$.}
  Assign $p \leftarrow K (K + 1) / 2$ and $q \leftarrow K (K - 1) / 2$.\;
  Compute $\hat{Z}$, the adjacency spectral embedding of $A$ using signature
  $(p, q)$.\;
  Compute $\hat{P} = \hat{Z} I_{pq} \hat{Z}^\top$, the estimate of the edge 
  probability matrix based on the adjacency spectral embedding.\;
  Arrange the rows and columns of $\hat{P}$ by cluster such that each 
  $\hat{P}^{(kl)}$ block consists of estimated edge probabilities between 
  clusters $k$ and $l$.\;
  \For{$k, l = 1, ..., K$, $k \leq l$} {
    Compute $\hat{P}^{(kl)} = U \Sigma V^\top$, the SVD of the $kl^{th}$ 
    block.\;
    Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$. 
    Assign $\sigma^2 \leftarrow \Sigma_{11}$.\;
    Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\hat{\lambda}^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

**Theorem 5**. Under regularity and sparsity assumptions, and under the further 
assumption that $K$ is fixed,

@@

*Proof (sketch)*.  
Let $P$ and $A$ be organized by cluster such that the elements of blocks 
$P^{(kl)}$ and $A^{(kl)}$ correspond to edges between clusters $k$ and $l$. 
By definition, $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. This implies 
that $P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top$ has one singular value 
and one pair of singular vectors. By convention, we can let 
$\lambda^{(kl)} = \sigma_{kl} u^{(kl)}$. Let 
$\hat{U}^{(kl)} \hat{\Sigma}^{(kl)} (\hat{V}^{kl})^\top$ be the singular value 
decomposition of $A^{(kl)}$, and let 
$\hat{\sigma}_{kl} \hat{u}^{(kl)} (\hat{v}^{(kl)})^\top$ be its one-dimensional 
approximation. Then 
$\hat{\lambda}^{(kl)} = \hat{\sigma}_{kl} \hat{u}^{(kl)}$.[^alg4] Thus 
$\hat{\lambda}^{(kl)}$ is a one-dimensional adjacency embedding approximation 
of $\lambda^{(kl)}$.  
Then by Theorem 5 from \citet{rubindelanchy2017statistical}, the adjacency 
embedding $\hat{\lambda}^{(kl)}$ approximates $\lambda^{(kl)}$ at rate 
$\frac{(\log n_k)^c}{n_k^{1/2}}$.

[^alg4]: Note that this estimator is similar to but slightly worse than the 
estimator from Algorithm 4.

@@

For the real data example, we applied Algorithm 2 to the Leeds Butterfly 
dataset \cite{Wang_2018} consisting of visual similarity measurements among 832 
butterflies across 10 species. The graph was modified to match the example 
from \citet{noroozi2019estimation}: Only the 4 most frequent species were 
considered, and the similarities were discretized to $\{0, 1\}$ via 
thresholding. Fig. \ref{fig:butterfly} shows a sorted adjacency matrix sorted 
by the resultant the clustering.

@@

>>>>

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation"
author: John Koo, Minh Tang, Michael Trosset
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
fontsize: 12pt
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
bibliography: misc.bib
abstract: |
  In this paper, we connect two probabilistic models for graphs, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  clustering and parameter estimation. In particular, we note that the PABM 
  can be represented as latent positions such that points within the same 
  cluster lie on a subspace, and the subspaces that represent each 
  cluster are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding (ASE) of the GRDPG, we 
  are able to establish theoretical asymptotic results of our clustering and 
  parameter estimation methods for the PABM. 
---

@@

\newcommand{\diag}{\text{diag}}
\newcommand{\tr}{\text{Tr}}

@@

Statistical analysis on graphs or networks often involves the partitioning of 
a graph into disconnected subgraphs or clusters. This is often motivated by the 
belief that there exist underlying and unobserved communities to which each 
vertex of the graph belongs, and edges between pairs of vertices are determined 
by drawing from a probability distribution based on the community relationships 
between each pair. The goal of this analysis then is population community 
detection, or the recovery of the true underlying community labels for each 
vertex, up to permutation (with some additional parameter estimation being of 
possible interest), assuming some probabilistic graph model. One such model is 
the Stochastic Block Model (SBM), which assumes that the edge probability from 
one vertex to another follows a Beronulli distribution with fixed probabilities 
for each pair of community labels. The Popularity Adjusted Block Model (PABM) 
was then introduced by \citet*{307cbeb9b1be48299388437423d94bf1} as a 
generalization of the SBM to address the heterogeneity of edge probabilities 
within and between communities while still maintaining the community structure.

The Random Dot Product Graph (RDPG) model \cite{athreya2017statistical} is 
another graph model with Beronulli edge probabilities. Under this model, each 
vertex of the graph can be represented by a point in a latent Euclidean space 
such that the edge probability between any pair of vertices is given by their 
corresponding dot product in the latent space. The SBM is equivalent to a 
special case of the RDPG model in which all vertices of a given community share 
the same position in the latent space. It has also been shown that similar 
probabilistic graph models, such as the Mixed Membership Stochastic Block 
Model, can be represented in this way \cite{rubindelanchy2017statistical}. An 
analogous property exists for the PABM, not under the RDPG model but under the *Generalized* Random Dot Product Graph (GRDPG) model. This relationship will be 
explored in this paper and exploited to construct algorithms for community 
detection and parameter estimation for the PABM.

In this paper, we will only consider undirected graphs, that is the edge weight 
from vertex $i$ to vertex $j$ is equal to the edge weight in the opposite 
direction, from vertex $j$ to vertex $i$. Furthermore, we will only consider 
unweighted graphs with binary ($0, 1$) edge weights We will also assume that 
graphs are hollow, i.e., there are no edges from a vertex to itself.

@@

## The popularity adjusted block model (PABM) \cite{noroozi2019estimation} and the generalized random dot product graph \cite{rubindelanchy2017statistical}

**Definition 1**. 
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ 
vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
$P^{(kl})$ describes the edge probabilities between vertices in communities 
$k$ and $l$. 
Let graph $G = (V, E)$ be an undirected, unweighted graph such 
that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
realization of $Bernoulli(P)$, i.e., 
$A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ for $i > j$ 
($A_{ij} = A_{ji}$ and $A_{ii} = 0$).

@@

**Definition 2**.
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ vertices, $V$. If $\exists X \in \mathbb{R}^{n \times d}$ such that 
$P = X I_{pq} X^\top$ for some $d, p, q \in \mathbb{N}$ and $p + q = d$, then 
graph $G = (V, E)$ with adjacency matrix $A$ such that $A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ for $i > j$ ($A_{ij} = A_{ji}$ and $A_{ii} = 0$) is a draw from the generalized random dot product graph model with latent positions $X$ and signature $(p, q)$. More precisely, if vertices $i$ and $j$ have latent positions $x_i$ and $x_j$ respectively, then the edge probability between the two is $P_{ij} = x_i^\top I_{pq} x_j$, and $X$ contains the latent positions as rows $x_i^\top$.

@@

**Definition 3**. The indefinite orthogonal group with signature $(p, q)$ is 
the set $\{Q \in \mathbb{R}^{d \times d} : Q I_{pq} Q^{\top} = I_{pq}\}$, 
denoted as $\mathbb{O}(p, q)$ \cite{rubindelanchy2017statistical}.

**Remark**. 
Like the RDPG, the latent positions of a GRDPG are not unique 
\cite{rubindelanchy2017statistical}. 
More specifically, if $P_{ij} = x_i^\top I_{pq} x_j$, then we also have for any 
$Q \in \mathbb{O}(p, q)$, 
$(Q x_i)^\top I_{pq} (Q x_j) = x_i^\top (Q^\top I_{pq} Q) x_j = 
x_i^\top I_{pq} x_j = P_{ij}$. 
Unlike in the RDPG case, transforming the latent positions by multiplication 
with $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or 
distances.

@@

* Positions 1, 5, 9 are fixed.
* The cycles of order 2 are $(2, 4)$, $(3, 7)$, and $(6, 8)$.

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation using the adjacency matrix.}
  \KwData{Adjacency matrix $A$, cluster assignments $1, ..., K$}
  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(kl)}\}_K$.}
  Arrange the rows and columns of $A$ by cluster such that each 
  $A^{(kl)}$ block consists of estimated edge probabilities between 
  clusters $k$ and $l$.\;
  \For{$k, l = 1, ..., K$, $k \leq l$} {
    Compute $A^{(kl)} = U \Sigma V^\top$, the SVD of the $kl^{th}$ 
    block.\;
    Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$. 
    Assign $\sigma^2 \leftarrow \Sigma_{11}$.\;
    Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\hat{\lambda}^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

**Theorem 5**. Under regularity and sparsity assumptions, and under the further 
assumption that $K$ is fixed and community labels are known,

@@

*Proof (sketch)*. 
Let $P$ and $A$ be organized by community such that the elements of blocks 
$P^{(kl)}$ and $A^{(kl)}$ correspond to the edges between clusters $k$ and $l$.

* *Case $k = l$*. $P^{(kk)}$ and $A^{(kk)}$ represent within-community edge 
probabilities and edges for community $k$.  
By definition, 
$P^{(kk)} = \lambda^{(kk)} (\lambda^{(kk)})^\top$. This implies that the 
signular value decomposition 
$P^{(kk)} = \sigma_{kk}^2 u^{(kk)} (u^{(kk)})^\top$ has one singular value and 
one pair of singular vectors ($P^{(kk)}$ is symmetric, so the left and right 
singular vectors are identical). Then $\lambda^{(kk)} = \sigma_{kk} u^{(kk)}$.  
Let $\hat{U}^{(kk)} \hat{\Sigma}^{(kk)} (\hat{U}^{(kk)})^\top$ be the singular 
value decomposition of $A^{(kk)}$, and let 
$\hat{\sigma}_{kk}^2 \hat{u}^{(kk)} (\hat{u}^{(kk)})^\top$ be its 
one-dimensional approximation. Define 
$\hat{\lambda}^{(kk)} = \hat{\sigma}_{kk} \hat{u}^{(kk)}$. Then 
$\hat{\lambda}^{(kk)}$ is the adjacency spectral embedding approximation of 
$\lambda^{(kk)}$.  
Then by Theorem 5 from \citet{rubindelanchy2017statistical}, the adjacency 
spectral embedding $\hat{\lambda}^{(kk)}$ approximates $\lambda^{(kk)}$ at rate 
$\frac{(\log n_k)^c}{\sqrt{n_k}}$.

* *Case $k \neq l$*. $P^{(kl)}$ and $A^{(kl)}$ represent edge probabilities and 
edges between communities $k$ and $l$.  
Let $R^{(kl)} = P^{(kl)} P^{(lk)}$. By definition, 
$P^{(kl)} P^{(lk)} = 
\lambda^{(kl)} (\lambda^{(lk)})^\top \lambda^{(lk)} (\lambda^{(kl)})^\top$. 
Noting that the two middle terms compose a scalar, we can rewrite 
$R^{(kk)} = c \lambda^{(kl)} (\lambda^{(kl)})^\top$, where $c = \tr(P^{(lk)})$. 
As in the $k = l$ case, we note that the singular value decomposition 
$R^{(kl)} = \sigma_{kl}^2 u^{(kl)} (u^{(kl)})^\top$ is one-dimensional and 
$\lambda^{(kl)} = \sigma_{kl} u^{(kl)} / c$.  
Let $B^{(kl)} = A^{(kl)} A^{(lk)}$, and let 
$\hat{U}^{(kl)} \hat{\Sigma}^{(kl)} (\hat{U}^{(kl)})^\top$ be the singular 
value decomposition of $B^{(kl)}$. Let 
$\hat{\sigma}_{kl}^2 \hat{u}^{kl} (\hat{u}^{(kl)})^\top$ be the one-dimensional 
approximation of $B^{(kl)}$ based on the singular value decomposition. Define 
$\hat{\lambda}^{(kl)} = \hat{\sigma}_{kl} \hat{u}^{(kl)} / \tr(A^{(lk)})$. Then 
$\hat{\lambda}^{(kl)}$ is the adjacency spectral embedding approximation of 
$\lambda^{(kl)}$.  
Then by Theorem 5 from \citet{rubindelanchy2017statistical}, the adjacency 
spectral embedding $\hat{\lambda}^{(kl)}$ approximates $\lambda^{(kl)}$ at rate 
$\frac{(\log n_k)^c}{\sqrt{n_k}}$.

@@

In the first real data example, we applied Algorithm 2 to the Leeds Butterfly 
dataset \cite{Wang_2018} consisting of visual similarity measurements among 832 
butterflies across 10 species. The graph was modified to match the example 
from \citet{noroozi2019estimation}: Only the 4 most frequent species were 
considered, and the similarities were discretized to $\{0, 1\}$ via 
thresholding. Fig. \ref{fig:butterfly} shows a sorted adjacency matrix sorted 
by the resultant clustering.

@@

```{r}
# this portion is from the code provided by Sengupta and Chen

setwd('~/Downloads/codes_and_data')
library("igraph")
library("clue")

##### Data input and pre-processing #####
foo=read.table("BritishMPcomm.txt",header=T)
mat=read.table("politicsuk-retweets.mtx",header=F)
mat = mat[-1,]
A = matrix(0,nrow(foo),nrow(foo))
for (k in 1:nrow(mat)){ #adjacency matrix with 5 parties
  edgei = mat[k,1]
  edgej = mat[k,2]
  i = which(foo[,1]==edgei)
  j = which(foo[,1]==edgej)
  A[i,j]<-1;A[j,i]<-1
}
b.true <- foo[,2]
levels(b.true) <- 1:5
b.true <- as.numeric(b.true)
nodes1 = which(b.true==1); nodes2 = which(b.true==2)
nodes = c(nodes1, nodes2)
A <- A[nodes,nodes] # Select only the two big communities
G<-graph.adjacency(A, mode="undirected")
foo1 <- clusters(G, mode="strong") 
nodes2 <- which(foo1$membership==1) # largest connected component (basically 31 nodes are disconnected)
A <- A[nodes2,nodes2] # form network with 329 nodes
b.true <- b.true[nodes2]
```

```{r}
zhat <- cluster.pabm(A, 2, use.all = TRUE, normalize = FALSE)
clust.acc <- mean(zhat == b.true)
```

```{r mp, cache = TRUE, fig.width = 2, fig.height = 2, fig.cap = 'Adjacency matrix of the British MPs dataset after sorting by the clustering outputted by Algorithm 2.'}
plot.A(A[order(zhat), order(zhat)])
```

In the second example, we applied Algorithm 2 to the British MPs Twitter 
network \cite{greene2013producing}. The original data consists of 419 vertices, 
each corresponding to a British MP belonging to one of five political parties. 
For this data analysis, we subsetted the data as described by 
\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the same 
dataset, resulting in 329 vertices and two parties ($n = 329, K = 2$). 
\citet{307cbeb9b1be48299388437423d94bf1} were able to obtain a classification 
accuracy of `r round(328 / 329 * 100, 1)`\%. Algorithm 2 obtained a lower 
classification accuracy of `r round(clust.acc * 100, 1)`\%.
